\myTitle

### 2015-03-11 tex/main-body-text.tex
Eric edited the manuscript.  Pretty close to ready to go now.

However, here we identify 
the close relationship between {\sc flock}'s algorithm and the program
{\sc structure}.  Although the algorithm in {\sc flock}
is, as the authors point out, deterministic (apart from a random initial allocation of 
individuals to clusters), it can be interpreted as a 
limiting case of a probabilistic search algorithm applied to one of {\sc structure}'s
models.  More specifically, the {\sc flock} 
algorithm is a special case of the simulated annealing
algorithm for finding the Bayesian maximum-{\em a-posteriori}
estimate from a marginalized form of the {\sc structure} model with no
admixture and non-correlated allele frequencies.

Being able to interpret {\sc flock} in this fashion should help users
interpret output from {\sc flock} and predict when it might and might not
give substantially different results than {\sc structure}. Below we describe
the correspondence between the programs in more detail, and then, for illustration,
create the progam {\sc flockture} by adapting the code of {\sc structure} to 
use a simulated annealing approach. We then compare results from large simulated 
datasets with varying degrees of population differentiation.

@@

The first step of the flock algorithm is initialization, during which 
the program randomly allocates each of the $N$ individuals to one of 
$K$ clusters.  Then a number of reallocation steps are performed.  During every one of 
these steps, each individual is given the chance to be reallocated 
(\ie moved to a different cluster).  This is done on the basis of 
maximum likelihood: as \citeauthor{Duc&Tur2009} say (2009, p.~1335), ``Re-allocations are 
performed following multilocus maximum likelihood (Paetkau et al. 1995).''

From that description, it is not clear how \citeauthor{Duc&Tur2009} treat 
alleles that appear in the focal
individual but do not appear within any other individuals within a cluster.
The original approach of \citet{Paetkauetal1995} merely added 0.01 or some other
small value to each allele frequency that was 0, and then renormalized 
the allele frequencies to sum to 1.  In later works, Paetkau and colleagues
also employed the ``Bayesian'' approach of \citet{Ran&Mou1997} in their 
software {\sc geneclass2}  \citep{Piryetal2004}.  If \citeauthor{Duc&Tur2009}
use the approach of \citet{Ran&Mou1997}, then the $i\thh$ individual will
be reallocated to whichever cluster gives the highest value to 
$P(Y_{i}~|~Z_i=k,~Z_{(-i)},~Y_{(-i)})$, which is 
exactly the probability
defined in (\ref{eq:cdm-multi-prob}).  
If \citeauthor{Duc&Tur2009} use the simpler ``add 0.01 and renormalize''
formulation of \citet{Paetkauetal1995}, then their reallocations are based
on maximizing a likelihood that, although it may not be formally identical to
the probability defined in (\ref{eq:cdm-multi-prob}), is nearly identical to it.

@@

\subsection*{The {\sc flock} algorithm and {\sc structure}}
The goal of {\sc structure}'s MCMC algorithm is to sample from the space of 
$Z$'s in proportion to their posterior probability.  If, all that was desired
was a point estimate of the $Z_i$'s that maximized the posterior probability,
then one standard approach to finding that maximum-{\em a-posteriori}, or MAP, 
estimate, would be to use simulated annealing \citep{Kirkpatricketal1983}.
To do simulated annealing in the {\sc structure} model with no admixture and
no correlated allele frequency prior, after integrating out the unknown
allele frequencies, the updates for each $Z_i$ would be made from the 
distribution
\[
\biggl[P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)})\biggr]^\beta
\]
which is the same full conditional distribution as the marginalized {\sc structure}
model (equation~\ref{eq:fc}), raised
to the power $\beta$.  In typical applications of simulated annealing,
$\beta$ is set to a starting value less than 1, and, as the algorithm proceeds,
$\beta$ is increased according to a ``cooling schedule''\citep{Hajek1988}, until
the algorithm finds a local maximum in the posterior probability surface.  
When $\beta$ is small, the process can easily traverse ``valleys''
in the posterior probability surface and has a better chance of searching
more of the space for the maximum value.  As $\beta$ gets larger, it becomes
more probable that the process will
move uphill on the probability surface, tending toward the local
maximum near where it currently is.

It should now be clear that the algorithm in {\sc flock} is a simulated annealing 
algorithm for the {\sc structure} model in which the cooling schedule is
``start with $\beta\rightarrow\infty$ and leave it there for the duration of 
the algorithm.''  As such, we can expect that it may be more susceptible 
to becoming trapped in local modes of the posterior probability. 
In the following section we show that {\sc structure} and {\sc flock} find similar
solutions in the analysis of a large microsatellite data set, though, as expected,
{\sc flock} seems to find the same solution less consistently than does {\sc structure}.

\section*{Comparison of {\sc structure} and {\sc flock}}
{\sc flock}, initially designed to map admixture in hybrid populations 
\citep{Duc&Tur2009}, was recently proposed as an alternative to 
{\sc structure} for estimating the number of populations from a collection with 
unsampled source populations \citep{Duc&Tur2012}. In doing so, 
{\sc flock} relies on individuals with like genotypes clustering together 
through an iterative process. It stands to reason then, a reasonable way 
to gauge how well the algorithm accomplishes the main objective 
of clustering individuals into their correct groups is to simulate data of known 
population structure with varying levels of differentiation and 
evaluate how well it does assuming the correct number of populations. The 
algorithme performance can then be measured with a zero-one 
loss function (where a loss of 1 is incurred if an individual is misassigned).

The implementation of {\sc flock} as a VBA procedure (macro) in 
Microsoft Excel precludes a rigorous evaluation of its behavior due 
to the inability to efficiently batch process. Consequenty, we adapted 
the source code from {\sc structure} as describe above to create {\sc flockture}
(code provided on GitHub repository https://github.com/eriqande/flockture) 
in order to accurately assess the strengths and weaknesses of the {\sc flock} 
algorithm. First, using simulated data we demonstrated that {\sc flock} 
and {\sc flockture} are essentially equivalent programs.
Second, we simulated and analyzed datasets with {\sc flockture} 
and three models in {\sc structure} (the admixture and correlated 
allele frequency (A.C) model, the no-admixture and correlated 
allele frequency (NA.C) model, and the no-admixture and non-correlated 
allele frequency (NA.NC) model) to help better understand
under what conditions, if any, running {\sc flock} may be preferable to {\sc structure} .

For each set of runs, {\sc flock} reports the 
normalized likelihood values (the exponentiated log-likelihood value for an individual's 
membership to each of $K$ clusters, scaled to sum to one) for a run belonging to the 
longest plateau - refered to as the 'best run' by \citep{Duc&Tur2012} (see below).
These values can be compared to the $q_i$ values
returned by {\sc structure}, which, in the no-admixture case, are posterior 
probabilities of cluster membership. \citet{Duc&Tur2012} describe a method 
for using {\sc flock} to estimate, $K$, the number of
subpopulations. The user must specify both the number of runs 
(different initial allocations of individuals 
to each cluster) as well as the number of iterations. (In each iteration every individual 
has the opportunity to be reallocated 
to a different cluster.) At the conclusion of each run, the log-likelihood value for 
membership to all $K$ clusters is computed
and the log likelihood difference (LLOD), the difference between the 
cluster with the highest likelihood 
and second highest likelihood, is calculated for each individual. {\sc flock} 
identifies runs that have converged to the same solution by detecting 
that they have the same mean LLOD score, and, in 
{\sc flock} parlance, a series of runs that have converged to the same 
solution are termed a ``plateau.'' The plateau record is a list of 
how many runs converged to the same allocation
of individuals. \citet{Duc&Tur2012} developed two stopping 
rules that they use to estimate $K$. \citet{Duc&Tur2012} advocate 
running  {\sc flock} until either a single plateau of
$\geq 6$ runs is reached or four consecutive $K$ values yield no 
plateaus. If the first stopping condition is satisfied, the $K$ for which 
there is a single plateau of $\geq 6$ runs is the point estimate for $K$. 
Alternatively, if multiple plateaus are observed with at least one plateau 
of $\geq 6$ runs, the largest value of $K$ that satisfies 
this conditon should be taken as the lower bound estimate of $K$. 
If no plateaus of $\geq 6$ runs are observed, $K$
is declared {\em undecided}.

To compare {\sc flock} and {\sc flockture} 4 datasets composed of 5 populations
were simulated using a coalescent based program (code provided on GitHub repository 
https://github.com/eriqande/flock-comment/simdata). Each of the 5 populations
were composed of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes
and an average pairwise $F_{ST}$ of 0.03, 0.02, 0.01, and 0.005 among
populations. Both programs were run nine times at $K=5$ with the default setting for all 
run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
LLOD threshold = 0). We judged {\sc flockture} as an accurate 
representation of {\sc flock} based on the
following criteria: (1) Clustering of individuals into corrected population as
 defined by the zero-one loss function, (2) individual $q_i$ values, and (3) plateau records. 
All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB of 
RAM running 64-bit Windows 7.

Once we were confident that no major differences in inference existed between 
{\sc flock} and {\sc flockture} we used {\sc flockture} to analyze 48 datasets for two marker types 
(microsatellites and single nucleotide polymorphisms (SNPs))  that encompassed a broader 
range of population differentiation. The 96 datasets were simulated as above (five populations 
composed of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes for the microsatellites
and 96 loci for the SNPs)
with average pairwise $F_{ST}$ values ranging between 0.004 and 0.045. 
 {\sc flockture} was run with the default run parameters and each run of 
{\sc structure} consisted of a 5,000 sweeep burn-in followed by a 20,000 sweep sample
from the posterior distribution. All runs were performed on a Macbook with a 2.3 GHz i7 processor
and 16GB of RAM running OS X ver. 10.7.5. The program {\sc clumpp} \citep{Jak&Ros2007} 
was used to relabel clusters among different runs,
while the program {\sc district} \citep{Rosenberg2004} was used to plot individual \textit{$q_i$} values.

\subsection*{{\sc flock} and {\sc flockture}} 
{\sc flock} and {\sc flockture} provided similar clustering of individuals among all four 
simulated datasets. The zero-one loss across all datasets had eqivalent results (Fig.~\ref{fig:FvFloss}).
When genetic differentiation is high both programs correctly apportion the majority of individuals and 
as genetic differentiation degrades among populations both programs show corresponding declines in 
assignment ability. Neither the mean nor range of loss between {\sc flock} and {\sc flockture} differs
substantially among the simulated datasets. These assignments were dependent on the normalized likelihood
values ($q_i$), which while they were comparable may be slightly biased toward the bounds (0,1)
for {\sc flock} (Fig.~\ref{fig:FvFqi}). In an attempt to understand the 
deviations being observed we took a subset of the 
first dataset (29 individuals from population 1 and 21 individuals from population 2) and initiated 
{\sc flock} with all individuals coming from population 1 starting in reference population 1 and all 
individuals from population 2 starting in reference population 2. {\sc flock} was run a second time
with the same starting clusters, but with the reference population labels switched. While the final allocation
of individuals matched the mean LLOD was slightly different (5.045009 vs. 5.028084). As a result, comparisons
between {\sc flock} and {\sc flockture} were limited to random initial starting points and should not 
be identical, but should have displayed equivalent behavior.

@@

\begin{figure*}
\centering
\begin{minipage}{.45\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{images/Figures-Pat/FlockvFlocktureLoss.pdf}
  \captionof{figure}{Zero-one loss function for {\sc flock} and {\sc flockture} for 9 replicate runs of simulated datasets comprised 
of five populations with varying levels of population differentiation:  $F_{ST} = $ (a) 0.03, (b) 0.02, (c) 0.01, and (d) 0.005.}
  \label{fig:FvFloss}
\end{minipage}%
\hfill
\begin{minipage}{.45\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{images/Figures-Pat/FlockvFlockture_qi.pdf}
  \captionof{figure}{Normalized likelihood values for {\sc flock} and {\sc flockture} for a single run of simulated
data with average pairwise $F_{ST} = $ 0.03. }
  \label{fig:FvFqi}
\end{minipage}
\end{figure*}

Run times for both programs were relatively short, albeit in the case of {\sc flock} 
non-trivial. {\sc flock} averaged 1966.67 ($\pm 64.32$) seconds
over all runs, while {\sc flockture} averaged 9.83 ($\pm 0.21$) seconds. Run times showed a 
general increase as population structure became more subtle, however {\sc flock} did not 
demonstrate a linear increase.

\subsection*{{\sc flockture} and {\sc structure}} 
As expected {\sc flockture} and {\sc structure} both perform extremely well on datasets with
well differentiated populations. Infrequently, at higher levels of population differentiation the 
NA.NC and NA.C models in {\sc structure} exhibited large losses (Fig.~\ref{fig:uSatLoss})
resulting from the assignment of individuals from two groups into a single cluster while 
individuals from another group were separated into two clusters
with intermediate $q_i$ values (Fig.~\ref{fig:FvSdistruct}  A). This phenomena 
(only present in a single microsatellite dataset) disappeared as population differentiation 
declined. As the subtlety of genetic structure 
increased loss for both programs increased and differences in the performance of
each model with different marker types emerged.
For the microsatellite dataset we observed that both programs
produce roughly equivalent results at higher levels of population differentiation. At intermediate 
values, the variance of the loss incurred by each model increased and certain models did marginally
better in assignment. No model did consistently better than others, until $F_{ST}$ droped 
below 0.0136 after which the loss for {\sc flockture} was markedly larger as was the among run
variance. The drastic increase in variance among runs was also observed with the SNP datasets. 
 {\sc flockture} did, however, have consistenlty smaller loss than  {\sc structure} at high to intermediate 
values of $F_{ST}$. Once $F_{ST}$ fell below 0.016 {\sc structure} began correctly assigning a 
higher proportion of individuals.

The number and length of plateaus {\sc flockture} encountered quickly 
diminished as the amount of differentiation among populations decreased. For the microsatellite datasets
only one run at $F_{ST} = 0.0339$ produced a single plateau of length greater than six. The other eight 
replicate runs yielded plateau records varying in length from two to eight, all of which had at 
least one plateau larger than thirteen. When $F_{ST}$ was larger than 0.0337 no plateau record
contained a plateau of length six or greater. No plateaus were observed for any run when 
$F_{ST}$ was larger than 0.02187. Results from the SNP datasets
were concordant. No run of {\sc flockture} resulted in a single plateau of length greater than six.
Only simulated datasets with pairwise $F_{ST}$ values greater 0.0323 had plateau records with a 
plateau greater than six. Despite the lack of plateaus observed for many of the simulated datasets 
graphs of the normalized likelihood values from the {\sc flockture} run with the highest 
likelihood were suggestive of population structure.

Plots of the normalized likelihood values from {\sc flockture} can be useful in discerning the presence of 
structure.  In situations where only a single run of {\sc flockture} produced a plateau of length 2 
(Fig.~\ref{fig:FvSdistruct} B) the variation in assignment of individuals among runs was qualitatively indistinguisable
and results were highly concordant with those of {\sc structure}. As population differentiation degrades 
and plateaus are not observed, the run with the highest likelihood from {\sc flockture} still yields
similar clusters as {\sc structure} but with more variation among runs (Fig.~\ref{fig:FvSdistruct} C \& D).
As pairwise $F_{ST}$ decreases and the variation among {\sc flockture} runs increases the normalized likelihood
values for assignment remain high in contrast to the $q_i$ values in each of the three models of {\sc structure}.

@@

\section*{Conclusions}
In the past decade, the use of genetic markers to identify population structure and to attempt inference of
the number of genetic clusters ($K$) in a collection has dramatically increased in the fields of ecology, evolution, 
epidemiology, and conservation
genetics. The most commonly used program for doing this is {\sc structure} \citep{Pritchardetal2000,Falushetal2003}, 
a full-featured program with many options that is widely-used, well-tested, and universally regarded as a standard tool amongst molecular ecologists. 
Estimating $K$ is a difficult task, and because it is often based on individual assignment we 
evaluated these programs on their ability to correctly assign indivdiuals at the true value
of $K$. We have shown here that the model incorporated by {\sc flock} is essentially equivalent to the no-admixture model 
with non-correlated allele frequencies within the program {\sc structure}. 
We adapted the source code of {\sc structure} to create {\sc flockture} and show that
 {\sc flock} and {\sc flockture} are essentially equivalent programs. 
In comparing results from {\sc flockture} and {\sc structure} 
on simulated datasets with varying levels of differnetiation among populations
we find that the two programs give similar results.

Though we observed similar results
between the two programs at large values of $F_{ST}$, 
{\sc flockture}'s `best-run' solutions consistently assigned fewer individuals correctly
and became somewhat inconsistent at lower values of $F_{ST}$.
This is likely a consequence of the low population divergence and also
the large size of the search space over possible partitions. 
The likelihood surface defined on partitions of this space will often
be multimodal---possessing numerous local peaks and troughs.
The peak that  {\sc flockture}'s algorithm finds is entirely determined by the initial, random, allocation of 
individuals, as the algorithm proceeds deterministically after the initial random allocation.
By having a large number of individuals
that can be arranged into 5 groups defines a large partion space.
Decreasing $F_{ST}$ leads to more homogenous 
populations whose multilocus genotypes have a smaller difference in 
likelihood of assigning to any given cluster. This could act by increasing the number of
local peaks, leading {\sc flockture} to settle on many different solutions. The problem of getting 
caught in local modes is not unique to {\sc flock}---{\sc structure} also can converge to 
different solutions in complex spaces; however our results suggest this is a bigger problem for {\sc flockture}.
In species characterized by large subpopulations and low divergence, as is the case with many 
marine fish, the use of {\sc flock} should probably be avoided.

>>>>

Given the above sort of emphasis on the distinctiveness of the {\sc flock}
algorithm, it would not be surprising if most readers incorrectly believed {\sc flock}
to be based upon a completely different model than the well-known program
{\sc structure}. Here we remedy such confusion with a simple derivation
that identifies 
the close relationship between the {\sc flock} algorithm and the program
{\sc structure}.  Although the algorithm in {\sc flock}
is, as the authors point out, deterministic (apart from a random initial allocation of 
individuals to clusters), it can be interpreted as a 
limiting case of a probabilistic search algorithm applied to one of {\sc structure}'s
models.  More specifically, the {\sc flock} 
algorithm is a special case of the simulated annealing
algorithm for finding the Bayesian maximum-{\em a-posteriori}
estimate from a marginalized form of the {\sc structure} model with no
admixture and non-correlated allele frequencies.  Thus, while the models underlying
{\sc structure} and {\sc flock} are nearly indistinguishable, {\sc structure}'s inference
relies on MCMC samples from the posterior distribution, while
{\sc flock}'s inference involves finding a local maximum in the posterior surface.

Being able to interpret {\sc flock} in this fashion should help users
interpret output from {\sc flock} and predict when it might and might not
give substantially different results than {\sc structure}. Below we describe
the correspondence between the programs in more detail, and then, for illustration,
and to aid in application to simualted data, we
create the progam {\sc flockture} by adapting the code of {\sc structure} to 
use a simulated annealing approach. We then compare results of the two
programs from large 
datasets simulated with varying degrees of population differentiation.

@@

The first step of the {\sc flock} algorithm is initialization, during which 
the program randomly allocates each of the $N$ individuals to one of 
$K$ clusters.  Then a number of reallocation steps are performed.  During every one of 
these steps, each individual is given the chance to be reallocated 
(\ie moved to a different cluster).  This is done on the basis of 
maximum likelihood: as \citeauthor{Duc&Tur2009} say (2009, p.~1335), ``Re-allocations are 
performed following multilocus maximum likelihood (Paetkau et al. 1995).''

From that description, it is not clear how \citeauthor{Duc&Tur2009} treat 
alleles that appear in the focal
individual but do not appear within any other individuals within a cluster.
The original approach of \citet{Paetkauetal1995} merely added 0.01 or some other
small value to each allele frequency that was 0, and then renormalized 
the allele frequencies to sum to 1.  In later works, Paetkau and colleagues
also employed the ``Bayesian'' approach of \citet{Ran&Mou1997} in their 
software {\sc geneclass2}  \citep{Piryetal2004}.  If \citeauthor{Duc&Tur2009}
used the approach of \citet{Ran&Mou1997}, then the $i\thh$ individual will
be reallocated to whichever cluster gives the highest value to 
$P(Y_{i}~|~Z_i=k,~Z_{(-i)},~Y_{(-i)})$, which is 
exactly the probability
defined in (\ref{eq:cdm-multi-prob}).  
However, it turns out that \citeauthor{Duc&Tur2009} use the simpler ``add 0.01 and renormalize''
formulation of \citet{Paetkauetal1995}---though replacing $0.01$ with $1/(2n)$ where 
$n$ is the number of individuals curently allocated to cluster (Pierre Duchesne, pers. comm.)---so 
their reallocations are based
on maximizing a likelihood that, although it may not be formally identical to
the probability defined in (\ref{eq:cdm-multi-prob}), is nearly identical to it.

@@

\subsection*{The {\sc flock} algorithm and {\sc structure}}
The goal of {\sc structure}'s MCMC algorithm is to sample from the space of 
$Z$'s in proportion to their posterior probability.  If all that was desired
was a point estimate of the $Z_i$'s that maximized the posterior probability,
then one standard approach to finding that maximum-{\em a-posteriori}, or MAP, 
estimate, would be to use simulated annealing \citep{Kirkpatricketal1983}.
To do simulated annealing in the {\sc structure} model with no admixture and
no correlated allele frequency prior, after integrating out the unknown
allele frequencies, the updates for each $Z_i$ would be made from the 
distribution
\[
\biggl[P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)})\biggr]^\beta
\]
which is the same full conditional distribution as the marginalized {\sc structure}
model (equation~\ref{eq:fc}), raised
to the power $\beta$.  In typical applications of simulated annealing,
$\beta$ is set to a starting value less than 1, and, as the algorithm proceeds,
$\beta$ is increased according to a ``cooling schedule''\citep{Hajek1988}, until
the algorithm finds a local maximum in the posterior probability surface.  
When $\beta$ is small, the process can easily traverse ``valleys''
in the posterior probability surface and has a better chance of searching
more of the space for the maximum value.  As $\beta$ gets larger, it becomes
more probable that the process will
move uphill on the probability surface, tending toward the local
maximum near where it currently is.

It should now be clear that the algorithm in {\sc flock} is a simulated annealing 
algorithm for the {\sc structure} model in which the cooling schedule is
``start with $\beta\rightarrow\infty$ and leave it there for the duration of 
the algorithm.''  As such, we should not be surprised if it is prone 
to becoming trapped in local modes of the posterior probability, though
the practical implications of that are not yet clear.

\section*{Comparison of {\sc structure} and {\sc flock}}
{\sc flock}, initially designed to map admixture in hybrid populations 
\citep{Duc&Tur2009}, was recently proposed as an alternative to 
{\sc structure} for estimating the number of populations from a collection with 
unsampled source populations \citep{Duc&Tur2012}. Whatever the ultimate goal
of its application,  
{\sc flock} relies on clustering individuals with like genotypes together  
through an iterative process. Thus, a reasonable way 
to gauge how well the algorithm functions is to simulate data of known 
population structure with varying levels of differentiation and 
evaluate how well it does in recovering that structure when the data are
analyzed assuming
the number of subpopulations is exactly the number of subpopulations in the simulation. The 
algorithm performance can then be measured with a zero-one 
loss function (where a loss of 1 is incurred if an individual is group with the wrong cluster).

The implementation of {\sc flock} as a VBA procedure (macro) in 
Microsoft Excel retards rigorous evaluation of its behavior due 
the its slow execution speed and 
the inability to efficiently batch-process multiple data sets. Consequenty, we adapted 
the source code from {\sc structure} as described above to create {\sc flockture}
(code provided on GitHub repository https://github.com/eriqande/flockture) 
in order to accurately assess the strengths and weaknesses of the {\sc flock} 
algorithm. First, using simulated data we demonstrate that {\sc flock} 
and {\sc flockture} are essentially equivalent programs.
Second, we simulate and analyze datasets with {\sc flockture} 
and three models in {\sc structure} (the admixture and correlated 
allele frequency (A.C) model, the no-admixture and correlated 
allele frequency (NA.C) model, and the no-admixture and non-correlated 
allele frequency (NA.NC) model) to help better understand
under what conditions, if any, running {\sc flock} may be preferable to {\sc structure}.

For each set of runs, {\sc flock} reports the 
normalized likelihood values (the exponentiated log-likelihood value for an individual's 
membership to each of $K$ clusters, scaled to sum to one) for a run belonging to the 
longest plateau---refered to as the `best run' by \citep{Duc&Tur2012} (see below).
These values can be compared to the $q_i$ values
returned by {\sc structure}, which, in the no-admixture case, are posterior 
probabilities of cluster membership. In {\sc flock} the user must specify both the number of runs 
(different initial allocations of individuals 
to each cluster) as well as the number of iterations (in each iteration every individual 
has the opportunity to be reallocated 
to a different cluster).  At the conclusion of each run, the log-likelihood value for 
membership to all $K$ clusters is computed
and the log likelihood difference (LLOD), the difference between the 
cluster with the highest likelihood 
and second highest likelihood, is calculated for each individual. {\sc flock} 
identifies runs that have converged to the same solution by detecting 
that they have the same mean LLOD score, and, in 
{\sc flock} parlance, a series of runs that have converged to the same 
solution are termed a ``plateau.'' The plateau record is a list of 
how many runs converged to the same allocation
of individuals.

To compare {\sc flock} and {\sc flockture}, 4 datasets composed of 5 populations
were simulated using a coalescent based program (code provided on GitHub repository 
https://github.com/eriqande/flock-comment/simdata). The 5 populations
were composed of 250, 250, 400, 550, and 550 individuals, respectively, all typed  at 15 loci.
Each of the four data sets was simulated with a different 
average pairwise $F_{ST}$ between the populations: 0.03, 0.02, 0.01, and 0.005. 
Both {\sc flock} and {\sc flockture} were run nine times at $K=5$ with the default setting for all 
run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
LLOD threshold = 0). We judged {\sc flockture} as an accurate 
representation of {\sc flock} based on the
following criteria: (1) Clustering of individuals into corrected population as
 defined by the zero-one loss function, (2) individual $q_i$ values, and (3) plateau records. 
All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB of 
RAM running 64-bit Windows 7.

Once we were confident that no major differences in inference existed between 
{\sc flock} and {\sc flockture} we used {\sc flockture} to analyze 48 datasets for each of two marker types 
(microsatellites and single nucleotide polymorphisms (SNPs))  that encompassed a broader 
range of population differentiation. The 96 datasets were simulated as above (five populations 
composed of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes for the microsatellites
and 96 loci for the SNPs)
with average pairwise $F_{ST}$ values ranging between 0.004 and 0.045. 
 {\sc flockture} was run with the default run parameters and each run of 
{\sc structure} consisted of a 5,000 sweep burn-in followed by a 20,000 sweep sample
from the posterior distribution. All runs were performed on a Macbook with a 2.3 GHz i7 processor
and 16GB of RAM running OS X ver. 10.7.5. The program {\sc clumpp} \citep{Jak&Ros2007} 
was used to relabel clusters among different runs,
while the program {\sc distruct} \citep{Rosenberg2004} was used to plot individual \textit{$q_i$} values.

\subsection*{{\sc flock} and {\sc flockture}} 
{\sc flock} and {\sc flockture} provided similar clustering of individuals among all four 
simulated datasets. The zero-one loss across all datasets had eqivalent results (Fig.~\ref{fig:FvFloss}).
When genetic differentiation is high both programs correctly cluster the majority of individuals and 
as genetic differentiation declines among populations both programs show a corresponding deterioration in 
clustering ability. Neither the mean nor range of loss between {\sc flock} and {\sc flockture} differs
substantially among the simulated datasets.

We attempted for some time to develop {\sc flockture} to yield
completely identical results to those from {\sc flock}; however we did not
persist in those endeavors when we realized that {\sc flock} itself does
not provide consistent results when expected by symmetry.  Briefly, In an attempt to understand the 
we took a subset of the 
first dataset (29 individuals from population 1 and 21 individuals from population 2) and initiated 
{\sc flock} with all individuals originating from population 1 being placed in reference population 1 and all 
individuals from population 2 starting in reference population 2. {\sc flock} was run a second time
with the same starting clusters, but with the reference population labels switched. While the final allocation
of individuals matched, the mean LLOD was slightly different (5.045009 vs. 5.028084), indicating it would
be futile to write {\sc flockture} to yield formally identical results to {\sc flock}---{\sc flock} itself
doesn't yield identical results when it should. Therefore, comparisons
between {\sc flock} and {\sc flockture} were limited to random initial starting points and so will not 
be identical, but do display equivalent behavior.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=.9\linewidth]{images/Figures-Pat/FlockvFlocktureLoss.pdf}
  \caption{Zero-one loss function for {\sc flock} and {\sc flockture} for 9 replicate 
  runs of simulated datasets comprised of five populations with varying
  levels of population differentiation:  $F_{ST} = $ (a) 0.03, (b) 0.02, (c) 0.01, and (d) 0.005.}
  \label{fig:FvFloss}
\end{center}
\end{figure}

{\sc flockure} ran much more quickly than {\sc flock}.  {\sc flock} 
averaged 1966.67 ($\pm 64.32$) seconds per run,
over all runs, while {\sc flockture} averaged 9.83 ($\pm 0.21$) seconds---a 200-fold improvement in 
speed.

\subsection*{{\sc flockture} and {\sc structure}} 
As expected {\sc flockture} and {\sc structure} both perform extremely well on datasets with
well differentiated populations. Infrequently, at higher levels of population differentiation the 
NA.NC and NA.C models in {\sc structure} exhibited large losses (Fig.~\ref{fig:uSatLoss})
resulting from the assignment of individuals from two groups into a single cluster while 
individuals from another group were separated into two clusters
with intermediate $q_i$ values (Fig.~\ref{fig:FvSdistruct} A). This phenomenon 
(only present in a single microsatellite dataset) is an instance of {\sc structure}'s
Markov chain failing to converge well in the short runs we were using. As the subtlety of genetic structure 
increased loss for both programs increased and differences in the performance of
each model with different marker types emerged.
For the microsatellite dataset we observed that both programs
produce roughly equivalent results at higher levels of population differentiation. At intermediate 
values, the variance of the loss incurred by each model increased and certain models did marginally
better in assignment. No model did consistently better than others, until $F_{ST}$ dropped 
below 0.0136 after which the loss for {\sc flockture} was markedly larger as was the among run
variance. The drastic increase in variance among runs was also observed with the SNP datasets. 
{\sc flockture} did, however, have consistently smaller loss than  {\sc structure} at high to intermediate 
values of $F_{ST}$. Once $F_{ST}$ fell below 0.016, however, {\sc structure} began correctly assigning a 
higher proportion of individuals.

The number and length of plateaus {\sc flockture} encountered quickly 
diminished as the amount of differentiation among populations decreased. For the microsatellite datasets,
when $F_{ST}$ was less than 0.0337, no plateau record
contained a plateau of length six or greater. No plateaus were observed 
(i.e. no two runs of {\sc flockture} gave exactly the same solution) when 
$F_{ST}$ was smaller than 0.02187. Results from the SNP datasets
were concordant.
Only simulated datasets with pairwise $F_{ST}$ values greater 0.0323 had plateau records with a 
plateau greater than six. Despite the lack of plateaus observed for many of the simulated datasets, 
graphs of the normalized likelihood values from the {\sc flockture} run with the highest 
likelihood were suggestive of population structure.

Plots of the normalized likelihood values from {\sc flockture} can be useful in discerning the presence of 
structure.  In situations where only a single run of {\sc flockture} produced a plateau of length 2 
(Fig.~\ref{fig:FvSdistruct} B) the variation in assignment of individuals among runs was qualitatively indistinguishable
and results were highly concordant with those of {\sc structure}. As population differentiation declines 
and plateaus are not observed, the run with the highest likelihood from {\sc flockture} still yields
clusters similar to those from {\sc structure} but with more variation among runs (Fig.~\ref{fig:FvSdistruct} C \& D).
As pairwise $F_{ST}$ decreases and the variation among {\sc flockture} runs increases, the normalized likelihood
values for assignment remain high in contrast to the $q_i$ values in each of the three models of {\sc structure}.

@@

\section*{Conclusions}
In the past decade, the use of genetic markers to identify population structure and to attempt inference of
the number of genetic clusters ($K$) in a collection has dramatically increased in the fields of ecology, evolution, 
epidemiology, and conservation
genetics. The most commonly used program for doing this is {\sc structure} \citep{Pritchardetal2000,Falushetal2003}, 
a full-featured program with many options that is widely-used, well-tested, and universally regarded as a standard tool amongst molecular ecologists.

We have shown here that the model incorporated by {\sc flock} is analytically equivalent to {\sc structure}'s 
no-admixture model 
with non-correlated allele frequencies.
We adapted the source code of {\sc structure} to create {\sc flockture} and show that
{\sc flock} and {\sc flockture}  perform equivalently. 
In comparing results from {\sc flockture} and {\sc structure} 
on simulated datasets with varying levels of differentiation among populations
we find that the two programs give similar results.

Though we observed similar results
between the two programs at large values of $F_{ST}$, 
{\sc flockture}'s `best-run' solutions consistently assigned fewer individuals correctly
and became somewhat inconsistent at lower values of $F_{ST}$.
This is likely a consequence of the low population divergence and also
the large size of the search space over possible partitions. 
The likelihood surface defined on partitions of this space will often
be multimodal---possessing numerous local peaks and troughs.
The peak that  {\sc flockture}'s algorithm finds is entirely determined by the initial, random, allocation of 
individuals, as the algorithm proceeds deterministically after the initial random allocation.
Decreasing $F_{ST}$ leads to more homogenous 
populations whose multilocus genotypes have a smaller difference in 
likelihood of assigning to any given cluster. This likely increases the number of
local peaks, leading {\sc flockture} to settle on many different solutions. The problem of getting 
caught in local modes is not unique to {\sc flock}---{\sc structure} also can converge to 
different solutions in complex spaces; however our simulationssuggest this is a more  
prevalent problem for {\sc flockture}.
{\sc flock} is not best-suited to species characterized by large
subpopulations and low divergence, as is the case with many 
marine fish.

One of the advantages of {\sc flock} over other methods for genetic clustering
is reported to be the short processing time for each run of {\sc flock}.
While \citet{Duc&Tur2012} observed faster processing times (relative to other methods), 
their comparision was based on setting {\sc flock} to perform 50 runs and 20 iterations for each 
$K$ starting at 2 and ending when one of the stopping conditions was met, and computing times for that analysis 
were compared to ten iterations of the admixture and correlated allele frequency model in {\sc structure} using 
a 50,000 sweep burn-in followed by a 200,000 sweep sample 
from the posterior for values of $K$ from 1 to three more than the true 
value of $K$. It is difficult to reliably assess differences in program run times when there is no clear consensus on what 
constitutes a reasonable comparison with respect to run length and number of runs, however, we show that if 
{\sc flock} were to be implemented using good coding practices in a fast language, rather than as
a VBA routine in Microsoft Excel,
it would be approximately 200 times faster. This is not an inconsequential difference and is one reason
we used {\sc flockture} for a large-scale simulation assessment of the {\sc flock} algorithm.
